Linear Stochastic Approximation- Constant Step-Size and Iterate Averaging

### 2017-05-10 intro.tex
intro backup before removing comments

A linear stochastic approximation (LSA) algorithm of the form given by

@@

where the iterates $\theta_t\in \R^d$ and step-sizes $\alpha_t>0$ are quantities related to the algorithm, while $b_t\in \R^d$ and $A_t\in \R^{\dcd}$ is the data sequence presented to the algorithm.
LSA algorithms are useful in machine learning and reinforcement learning (RL) applications, where the aim is to compute a $\ts\in \R^d$ using the noisy data sequence $(b_t,A_t)$. Specific examples include the stochastic gradient descent algorithm for the problem of linear predition with quadratic loss, and the temporal difference class (TD) of learning algorithms in RL.
An additional feature in these class of pplications is that $A_t$ turns out to be a rank-$1$ matrix and $A_t\theta_{t-1}$ can be obtained in $O(d)$, which is attractive due to the cheap per time step computational requirement.\par
It is of interest to understand the effect of the step-sizes (a quantity that can be controlled by the algorithm designer) on the performance of the LSA. The choice of the step-sizes $\alpha_t$ needs to trade-off learning rate versus filtering noise, so to speak. In general, one can guess that larger step-sizes might lead to faster learning and at the same time smaller step-sizes are better at handling the noise. A technique in literature that goes back to \cite{polyak,ruppert} is to use a constant step-size (CS) $\alpha_t=\alpha>0,\forall t\geq 0$  in conjunction with averaging of iterates , known as Polyak-Rupert (PR) averaging to filter the noise. The averaged iterates can be given by $\th_t\eqdef\frac{1}{t}\sum_{s=0}^t \theta_t$. Under the CS-PR technique $\theta_t$ is the internal vairable and $\thh_t$ is the output of the algorithm. Intuitively, the CS-PR technique can be seen to address the learning vs filtering trade-off, where the constant step-size takes care of the learning  and the averaging of the iterates takes care of the filtering. \par
Recently \cite{bach} considered stochastic gradient descent algorithm (SGD)\footnote{SGD is a special of the LSA algorithm in \eqref{eq:lsaintro}.} with CS-PR for the linear prediction problem with squared loss and $i.i.d$ sampling. A ``remarkable'' result by \cite{bach} is that there exists a constant step-size such that for any problem where the noisy data is bounded by a known constant, the expected squared prediction error after $t$ updates is at most $\frac{C}{t}$ with a constant $C>0$ that depends \emph{only} on the bound on the data.\todoc{Dimension?}
\paragraph{Motivation and Challenges:} In this paper, we wish to study the effect of CS-PR in the case of general LSA algorithms of form \eqref{eq:lsaintro}. Our study is motivated by the problem of linear value function approximation from experience replay in a batch setting in RL using TD class of algorithms. An important difference between the linear prediction problem considered by \cite{bach} and the general LSA setting we consider is that in the former has certain special strucutres and the latter does not. One of the special strucutre exploited by \cite{bach} is the fact that $A_t$ are real symmetric matrices. The general LSA setting does not enjoy such an additional structure presents an interesting challenge.\par
\paragraph{Contributions:} We now list our contributions in this paper
\item We show that in the case of general LSA, there exists instance dependent constant step size choice such that the mean squared error (MSE) of the CS-PR procedure given by $\EE{\norm{\thh_t-\ts}^2}$ is at most $O(\frac{1}{t})$. The MSE further can be broken down to a \emph{bias} term (arising due to the initial condition $\theta_0$) that decays at a rate $O(\frac{1}{t^2})$ and a \emph{variance} term (due to the noise) that decays at at rate $O(\frac{1}{t})$.
\item We observe that when the constant step-size is chosen too big it results in instability, however, a conservative choice of the constant step-size affects only the constants in our error bounds without affecting the asymptotic rate (which might not be true with other popular choices of step-sizes). We aruge heuristically that CS-PR procedure can be deployed in a pricinpled manner by using an outer loop that automatically reduces that constant step-size by detecting detects the blow up of iterates.

@@

 We only assume that $\EE{A_t}$ is Hurwitz, i.e., all its eigenvalues have positive real parts. This is the condition used in prior works by \cite{polyak-judisky,borkarbook}. Eventhough our proof technique resembles that by \cite{bach}, in addition, we make use of Jordon/Cyclic decomposition in a critical way to handle the general Hurwitz assumption. Our results are applicable to wide variety of TD algorithms such as TD(0), GTD/GDT2, TDC.
Further, the error dynamics has a simple interpretation in terms of behaviour of the sum of terms in a geometric series, and hence offers a simple and unified understanding of the various TD algroithms.

@@

A host of algorithms in machine learning and in particular, the temporal difference class of algorithms arising in reinforcement learning, turn out to be special cases of \eqref{eq:sa}, namley linear stochastic approximation (LSA) algorithms. LSA algorithms assume the general form

@@

\item The intuition behind the error analysis can be explained in simple terms when the noise free case is considered (i.e., $A_t=A$ in \eqref{eq:lsaintro}). In this case, the LSA is just a linear time invariant dynamical system where the error $\theta_t-\ts=(I-\alpha A)^t (\theta_0-\ts)$ and $\norm{\theta_t-\ts}=\rho^t\norm{\theta_0-\ts}$ (for some $0<\rho<1$). Now the PR-average is given by $\frac{\norm{\theta_t-\ts}^2}{t^2}=\frac{\sum_{s=0}^{t-1}\rho^s}{t^2} \norm{\theta_0-\ts}^2$. Note that the numerator is just sum of terms of a geomteric series, which is summable as long as the matrix $(I-\alpha A)$ has eigen values within the unit circle.
\item The fact that the setting covers even two timescale LSAs with constant step size, coupled with the intuitive interpretation of the error dynamics together present a simple and unified view of the various temporal difference learning algroithms in literature.
\item One of the major bottlenecks in selecting step-size for general LSA is the lack of problem knoweldge. One way to handle this issue is to choose a diminishing step size schedule such as $\alpha_t=\frac{1}{t}$. However, such a strategy can fail when the underlying matrices are ill-conditioned (see \Cref{sec:stepsizes}). Another popular approach has been to choose $\alpha_t=\frac{c}{c+t}$, where initially the step-sizes are held almost constant and then as $t$ increases they decay at rate $O(\frac{1}{t})$. While this approach can mitigate the ill-conditioning problem, it can at the same time amplify the effect of noise. In this paper, we argue informally that the constant step-size with PR-averaging of iterates can be a promising way to mitigate these issues, by periodically checking for possible blow up of the norm of the iterates and then reducing the step-size based on this information (see \Cref{sec:stepsizes}).

>>>>

A linear stochastic approximation (LSA) algorithm is given by

@@

where the iterates, $\theta_t\in \R^d$, and step-sizes, $\alpha_t>0$, are algorithmic quantities, and $(b_t,A_t)\in \R^d\times \R^{\dcd}$ is the noisy data sequence presented to the algorithm.
LSA algorithms are useful in machine learning and reinforcement learning (RL) applications, where the aim is to compute a $\ts\in \R^d$ using the noisy data sequence, using $O(d)$ updates per iteration. Some examples of LSAs include the stochastic gradient descent algorithm (SGD) for the problem of linear prediction with quadratic loss \cite{bach,bachaistats}, and the temporal difference class (TD) of learning algorithms in RL \cite{sutton,konda-tsitsiklis,gtd,gtd2,gtdmp}.\par
%An additional feature in these class of applications is that $A_t$ turns out to be a rank-$1$ matrix and $A_t\theta_{t-1}$ can be obtained in $O(d)$, which is attractive due to the cheap per time step computational requirement.\par
It is of interest to understand the effect of the step-sizes $\alpha_t$ (a quantity that can be controlled by the algorithm designer) on the performance of the LSA in \eqref{eq:lsaintro}. The choice of the step-sizes dictates learning rate, noise rejection and stability. Informally speaking, smaller step-sizes are better for noise rejection and larger step-sizes lead to faster learning, while at the same time too large step-sizes might result in instability of \eqref{eq:lsaintro}. Diminishing and square-summable step-sizes have been a popular choice for the step-sizes, where $\alpha_t\ra 0$ such that $\sum_{t\geq 0} \alpha_t=\infty$, $\sum_{t\geq 0}\alpha^2_t<\infty$ \cite{gtd2,borkarbook,konda-tsitsiklis}. An alternative technique, which we call LSA with constant step-size Polyak-Ruppert (CS-PR) averaging, is to run \eqref{eq:lsaintro} as a constant step-size LSA by choosing $\alpha_t=\alpha>0,\forall t\geq 0$, and output the average $\thh_t\eqdef\frac{1}{t}\sum_{s=0}^t \theta_t$. In LSA with CS-PR, $\theta_t$ is an internal variable and $\thh_t$ is the output of the algorithm.
Averaging in stochastic approximation to improve noise rejection is a procedure that goes back to  Polyak and Ruppert \cite{polyak-judisky,ruppert}.
%In what follows, when we say CS-PR for LSA or LSA with CS-PR, we allude to the case where \eqref{eq:lsaintro} is a constant step-size LSA with $\alpha_t=\alpha>0,\forall t\geq 0$, $\theta_t$ in \eqref{eq:lsaintro} is considered an internal variable and $\thh_t$ is the final output of the LSA with CS-PR.
%In LSA with CS-PR, \eqref{eq:lsaintro} is a constant step-size LSA with $\alpha_t=\alpha>0,\forall t\geq 0$, $\theta_t$ in \eqref{eq:lsaintro} is considered an internal variable and $\thh_t$ is the final output of the LSA with CS-PR.
Intuitively, the CS-PR technique can be seen to address the learning vs noise rejection trade-off, where the constant step-size takes care of the learning  and the averaging of the iterates takes care of the noise rejection. \par
Recently, \cite{bach} considered SGD\footnote{SGD is a special of the LSA algorithm in \eqref{eq:lsaintro}.} with CS-PR for the linear prediction problem with quadratic loss and $i.i.d$ sampling. A ``remarkable'' result by \cite{bach} is that there exists a constant step-size such that for any problem where the noisy data is bounded by a known constant, the expected squared prediction error after $t$ updates is at most $\frac{C}{t}$ with a constant $C>0$ that depends \emph{only} on the bound on the data.\par\todoc{Dimension?}
\paragraph{Motivation and Contribution:} In this paper, we wish to study LSA with CS-PR for $i.i.d$ data with bounded variance. An additional restriction we place is that the matrix $\EEP{A_t}$ (where $A_t\sim P$ ) is Hurwitz, i.e., all its eigenvalues have positive real parts. For a given $P$, we wish to measure the performance in terms of the mean square error (MSE) $\EEP{\norm{\thh_t-\ts}^2}$. Further, for a given class of distributions $\P$, we also would like to know whether it is possible to ensure uniform performance $\forall P\in \P$.
Our study is motivated by the problem \emph{policy evaluation} \cite{dann} using linear value function approximation from experience replay \cite{lin} in a batch setting \cite{lange} in RL using TD class \cite{sutton,konda-tsitsiklis,gtd,gtd2,gtdmp} of algorithms. We now list our contributions in this paper
%While asymptotically diminishing step-sizes are common in RL literature \cite{ilstd,gtdmp,korda-prashanth}, the constant step-size choice has also been used in the past \cite{gtd2}.
\begin{itemize}[leftmargin=*]%, before = \leavevmode\vspace{-\baselineskip}]
\item \textbf{Instance Dependent Bounds} We show that for a fixed $P$, there exists a $\alpha_P>0$ such that the mean squared error (MSE) of the LSA with CS-PR is at most $O(\frac{1}{t})$, whenever $\alpha\in (0,\alpha_P)$. The MSE further can be broken down to a \emph{bias} term (arising due to the initial condition $\theta_0$) that decays at a rate $O(\frac{1}{t^2})$ and a \emph{variance} term (due to the noise) that decays at at rate $O(\frac{1}{t})$.
\item \textbf{Uniform Step-Sizes:} We also show (via counter example) that there is a class $\P$ that does not `admit' a constant step-size choice that guarantees uniform performance $\forall P\in \P$. We also argue that the subclass $\P_{LR}$ related to SGD for linear regression does `admit' a problem independent constant step size, thereby making a connection between our framework and that considered by \cite{bach}.
\item \textbf{Automatic Step-Size:} We argue informally that for any given $P$, a constant step-size can be chosen in an `automated' fashion, and the performance of CS-PR is comparable to other popular step-size choices used for LSA in RL literature.

@@

The performance of CS-PR for general LSA has been previously studied under additive noise assumptions where $A_t=A$ for some fixed $A\in\R^{\dcd}$ and the noise is present only in $b_t$ \cite{polyak-judisky}. The setting we consider in this paper is known as the multiplicative noise setting, whose analysis requires tools that exploit the linear structure of the LSA \cite{bach}. However,
 in the absence of special structures (such as symmetry of $A_t$),  we make use of Jordan/Cyclic decomposition in a critical way to handle the general Hurwitz assumption. Our results are applicable to wide variety of TD algorithms such as TD(0), GTD/GDT2, TDC. Further, the error dynamics of the CS-PR technique for LSAs has a simple interpretation in terms of behaviour of the sum of terms in a geometric series, and hence offers a simple and unified understanding of its use in the various TD algorithms.
An important difference between the linear prediction problem considered by \cite{bach} and the general LSA setting we consider is that in the former has certain special structures and the latter does not. One of the special structure exploited by \cite{bach} is the fact that $A_t$ are real symmetric matrices. The general LSA setting does not enjoy such an additional structure presents an interesting challenge. Further, CS-PR for general LSAs has been studied previous only under the additive noise setting \cite{polyal-judisky}.\par
\paragraph{Organization:}

@@

A host of algorithms in machine learning and in particular, the temporal difference class of algorithms arising in reinforcement learning, turn out to be special cases of \eqref{eq:sa}, namely linear stochastic approximation (LSA) algorithms. LSA algorithms assume the general form

@@

\item The intuition behind the error analysis can be explained in simple terms when the noise free case is considered (i.e., $A_t=A$ in \eqref{eq:lsaintro}). In this case, the LSA is just a linear time invariant dynamical system where the error $\theta_t-\ts=(I-\alpha A)^t (\theta_0-\ts)$ and $\norm{\theta_t-\ts}=\rho^t\norm{\theta_0-\ts}$ (for some $0<\rho<1$). Now the PR-average is given by $\frac{\norm{\theta_t-\ts}^2}{t^2}=\frac{\sum_{s=0}^{t-1}\rho^s}{t^2} \norm{\theta_0-\ts}^2$. Note that the numerator is just sum of terms of a geometric series, which is summable as long as the matrix $(I-\alpha A)$ has eigen values within the unit circle.
\item The fact that the setting covers even two timescale LSAs with constant step size, coupled with the intuitive interpretation of the error dynamics together present a simple and unified view of the various temporal difference learning algorithms in literature.
\item One of the major bottlenecks in selecting step-size for general LSA is the lack of problem knowledge. One way to handle this issue is to choose a diminishing step size schedule such as $\alpha_t=\frac{1}{t}$. However, such a strategy can fail when the underlying matrices are ill-conditioned (see \Cref{sec:stepsizes}). Another popular approach has been to choose $\alpha_t=\frac{c}{c+t}$, where initially the step-sizes are held almost constant and then as $t$ increases they decay at rate $O(\frac{1}{t})$. While this approach can mitigate the ill-conditioning problem, it can at the same time amplify the effect of noise. In this paper, we argue informally that the constant step-size with PR-averaging of iterates can be a promising way to mitigate these issues, by periodically checking for possible blow up of the norm of the iterates and then reducing the step-size based on this information (see \Cref{sec:stepsizes}).
